E-graphs With Bindings

### 2025-01-18 LICS_2025/e_graphs_with_bindings/E-graphs-with-bindings.tex
string diagrams section

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest particularly for optimisation and synthesis using so called \emph{equality saturation}~\cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}, the repeated application of rewrites to an e-graph until a fixpoint is reached. 
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
Of particular interest to us is using e-graphs in the context of functional programming languages, which requires a representation of the $\lambda$-calculus using it. 
Pioneering work in this direction has been done by~\cite{koehler2022sketchguided}.
The well-known twin challenges of binding and capture-avoiding substitution remain however somewhat problematic in this work because DAN: IT IS UNCLEAR WHY THIS IS PROBLEMATIC IN LOC. CIT. THE TEXT BELOW IS UNCONVINCING. CAN YOU ILLUSTRATE THE PROBLEM WITH SOME SIMPLE EXAMPLES? 
\begin{quotation}
There are domains, however, where the effective usage of e-graphs requires \textit{ad hoc} tweaks to achieve the desired performance, namely domains that include variable \textit{binding} and the closely related notion of \textit{substitution}, for example, lambda calculus with $\beta$-reduction.
The first issue appears due to the fact that lambda calculus terms are quotiented by renaming of bound variables.
If such renaming is not accounted for by the language encoding, \textit{e.g.,} by using \textit{de Brujin} indices, populating an e-graph with $\alpha$-equivalent terms increases the size of an e-graph without adding any meaningful equivalences.
The other issue is that having binding entails having a substitution and computing substitution in e-graphs is hard as it is not a mere substitution of terms but rather a substitution of equivalence classes that represent a lot of terms at once.
\end{quotation}

@@

>>>>

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest particularly for optimisation and synthesis using so called \emph{equality saturation}~\cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}, the repeated application of rewrites to an e-graph until a fixpoint is reached. 
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
Of particular interest to us is using e-graphs in the context of functional programming languages, which requires a representation of the $\lambda$-calculus using it. 
Pioneering work in this direction has been done by~\cite{koehler2022sketchguided} where
the authors outlined the two interconnected challenges of binding and capture-avoiding substitution that arise when doing equality saturation for $\lambda$-calculus with e-graphs.
A key challenge is that as e-graphs represent multiple terms at once, with as much sharing as possible, using variable names for binding means that two $\alpha$-equivalent terms can not be shared as they are not identical.

\begin{red}
DRG: UNCLEAR. Even though such simple encoding makes expressing $\beta$-reduction as a rewrite rule in terms of explicit substitution simpler, applying such rules requires computing extra free variable predicates and populates a given e-graph with extra nodes for explicit substitution.
The commonly deployed solution to quotienting terms by $\alpha$-equivalence, encoding binding using $\textit{de Brujin}$ indices, still fails to represent $\alpha$-equivalent terms as being structurally equivalent for the purpose of e-graphs. 
Consider the simple example term $\lambda . \%0 ((\lambda . \%1 \%0) 42)$ the same binder is represented with $\%0$ and $\%1$, avoids explicit substitution when implementing $\beta$-reduction at the cost of including \textit{ad hoc} operators for index shifting into the rewrite rules.
Generally, the incurred redundancy these challenges remain problematic because variable binding is not inherent to e-graphs as they do not provide an internal support for it.
There are also other inchoate approaches to adding internal support for binding which, however, make the formulation of e-graph considerably more complicated~\cite{slotted-egraphs}.
\end{red}

@@

We build on the technology of string diagrams, which is by now well developed in a number of areas. 
String diagrams have the advantage of being a direct bridge from the language of category theory, which in turn can be used to give semantic models to a variety of programming languages and logics, to the concrete language of graphs, all based on solid mathematical foundations. 
To solve the new problems discuss above, of dealing with binding and capture avoiding substitution, we just follow the categorical breadcrumb trail.

The general background literature for string diagrams is summarised in~\cite{https://arxiv.org/abs/2305.08768}.
The particular context of string diagrams with bindings is given in~\cite{https://arxiv.org/abs/2305.18945}, while string diagrams for e-graphs without bindings are defined in~\cite{https://arxiv.org/abs/2406.15882}.

String diagrams provide a geometric or topological framework for understanding monoidal categories, representing objects as \emph{strings} and morphisms as \emph{nodes}, where input strings flow in and output strings flow out. 
Key structures in monoidal categories---such as the monoidal product, tensor unit, braiding, and dualities---are represented through operations like placing, weaving, or bending strings. 
This visual approach makes complex operations such as the trace intuitive and reveals insights sometimes obscured by symbolic notation. 
Originating from notations by Hotz~\cite{??} (for automata) and Penrose~\cite{??} (for tensor networks), string diagrams generalise to any monoidal category and even bicategories, offering a dual perspective to commutative diagrams through Poincaré duality.

As described above, string diagrams are commonly an informal yet rigorous two-dimensional syntax for the internal language of a category. 
They can be formalised by representing them as graphs or hypergraphs~\cite{https://dl.acm.org/doi/full/10.1145/3502719}, which has two further benefits. 
The first benefit is that the property of two string diagrams being equal if and only if they ``look the same'' can be spelled out as graph isomorphism, eliminating any possible ambiguity of interpretation. 
The second benefit is that equational reasoning can also be formalised as (categorical) graph rewriting, for instance double-pushout rewriting (DPO).

Finally, to represent diagrammatically two concepts essential to both programming languages (binding) and to e-graphs (equivalence classes) we rely on the concept of \emph{functorial boxes} as introduced by Melli\`es~\cite{https://link.springer.com/chapter/10.1007/11874683_1}. 
The diagrammatic boxes require the introduction of further structure to (hyper)graphs in the form of a hierarchical layering of the graphs, as studied in \emph{loc. cit.}.

DRG TODO: ADD AN EXAMPLE OF A STRING DIAGRAM HERE BEFORE DISCUSSING IT IN THE NEXT SECTIONS. BEST WE HAVE A RUNNING EXAMPLE THAT IS RELATED TO THE DIFFICULTY OF REPRESENTING E-GRAPHS FOR LAMBDA CALCULUS.
